|

### 2021-12-02 07-Chp-Triangulation.Rmd
Working version!
End of day:

### Qualitative triangulation {#qual-tri}

@@

As an illustrative example, the bias and indirectness inherent to the estimate of the effect of statin use on Alzheimer's disease in the CPRD are considered here. For this result, there were three domains at greater than low risk of bias: Bias due to confounding, Bias due to definition of the outcome, and Bias in the selection of the reported result (Figure \@ref(fig:biasDirectionSingle)). In Domain 1 (bias due to confounding), the study did not adjust for ApoE4 status, which is predicted to mask any true association between statins and Alzheimer's disease (see Section \@ref(cprd-limitations) for a fuller discussion). As such, the bias is predicted to _favour the comparator_ and so the absolute direction of bias is to the right. In Domain 7 (bias in measurement of the outcome), there was a high risk of differential misclassification on the basis of the exposure. Here, history of statin use may make a diagnosis of vascular dementia more likely than Alzheimer's disease, and so for this outcome, this bias is predicted to _favour the experiment_ (in this case, an intervention). As such, the absolute direction of bias is to the left. In Domain 7 (bias in selection of the reported result), for consistency with the other NRSI, in the absence of a protocol the bias is assumed to be _away from the null_. Given the position of the effect estimate below the null, the absolute direction of bias in this domain is to the left.

@@

@@

Ideally, these prior distributions would be based on empirical data on the effect of different domains of bias/indirectness on a result. Most meta-epidemiological studies examining the effect of different methodological issues studied only randomised controlled trials,[@amer2021; @page2016] and the estimates of bias they produce may not hold for non-randomised study designs. However, the prior distributions defined here are broadly comparable to previously reported estimates of the impact of bias in non-randomised trials. For example, a previous simulation study estimated that, in the absence of true effect of LDL-c, ApoE4 would induce a RR of 1.09 with risk of dementia per 1-SD increase in LDL-c.[@iwagami2021] This is comparable to the mean assigned to the prior distribution mapped to "Moderate" additive bias (N(0.08,0.05), Table \@ref(tab:priorsAdd-table)).

>>>>

### Qualitative triangulation / narrative comparison {#qual-tri}

@@

As an illustrative example, the bias and indirectness inherent to the estimate of the effect of statin use on Alzheimer's disease in the CPRD are considered here. For this result, there were three domains at greater than low risk of bias: Bias due to confounding, Bias due to definition of the outcome, and Bias in the selection of the reported result (Figure \@ref(fig:biasDirectionSingle)). In Domain 1 (bias due to confounding), the study did not adjust for _ApoE4_ status, which is predicted to mask any true association between statins and Alzheimer's disease (see Section \@ref(cprd-limitations) for a fuller discussion). As such, the bias is predicted to _favour the comparator_ and so the absolute direction of bias is to the right. In Domain 7 (bias in measurement of the outcome), there was a high risk of differential misclassification on the basis of the exposure. Here, history of statin use may make a diagnosis of vascular dementia more likely than Alzheimer's disease, and so for this outcome, this bias is predicted to _favour the experiment_ (in this case, an intervention). As such, the absolute direction of bias is to the left. In Domain 7 (bias in selection of the reported result), for consistency with the other NRSI, in the absence of a protocol the bias is assumed to be _away from the null_. Given the position of the effect estimate below the null, the absolute direction of bias in this domain is to the left.

@@

<!-- TODO Data-driven hypothesis testing - only choosing those the two strata that have some suggestion of an effect -->

@@

Ideally, these prior distributions would be based on empirical data on the effect of different domains of bias/indirectness on a result. Most meta-epidemiological studies examining the effect of different methodological issues studied only randomised controlled trials,[@amer2021; @page2016] and the estimates of bias they produce may not hold for non-randomised study designs. However, the prior distributions defined here are broadly comparable to previously reported estimates of the impact of bias in non-randomised trials. For example, a previous simulation study estimated that, in the absence of true effect of LDL-c, _ApoE4_ would induce a RR of 1.09 with risk of dementia per 1-SD increase in LDL-c.[@iwagami2021] This is comparable to the mean assigned to the prior distribution mapped to "Moderate" additive bias (N(0.08,0.05), Table \@ref(tab:priorsAdd-table)).

### 2021-12-02 08-Discussion.Rmd
End of day:

The first a large scale analysis of blood lipids in the CPRD, which found a slight raised risk of dementia in CPRD patients () with higher LDL-c measured at mid-life, driven by the finding for Alzheimer's disease.[@iwagami2021] These findings are  consistent with the findings presented in this thesis, particularly given the lack of adjustment for confounding by ApoE4. <!-- TODO The study says that simulated effect of ApoE4 is a 1.09 RR in the absence of a true causal effect. Main results are between 1.03 and 1.17. CI for most extreme results is (lower = 1.08) is not inconsistent with the effect introduced by ApoE4 which is just a single confounder out of many -->

@@

>>>>

The first a large scale analysis of blood lipids in the CPRD, which found a slight raised risk of dementia in CPRD patients () with higher LDL-c measured at mid-life, driven by the finding for Alzheimer's disease.[@iwagami2021] These findings are  consistent with the findings presented in this thesis, particularly given the lack of adjustment for confounding by ApoE4. <!-- TODO The study says that simulated effect of _ApoE4_ is a 1.09 RR in the absence of a true causal effect. Main results are between 1.03 and 1.17. CI for most extreme results is (lower = 1.08) is not inconsistent with the effect introduced by _ApoE4_ which is just a single confounder out of many -->

@@

<!-- TODO Incorporate Julian's comments below -->

<!-- REVIEW [JH] I think, on general reflection, that portraying this as an issue around “grey literature” may not be particularly useful. Preprints are arguably some of the least “grey” of the very ambiguously defined family of grey sources, such that some definitions of “grey” might not include them.
It seems to me that preprints are such a big and important part of the landscape now that you can just portray the methodology as dealing with preprints rather than including grey literature.
Interesting debate as to whether preprints are grey literature, given that many of them are increasingly becoming findable in electronic bibliographic databases. We have no clear and widely used definition of grey literature.  -->

### 2021-12-02 front-and-back-matter/_abstract.Rmd
End of day:

__Background__  
In the UK, an estimated 800000 people are currently living with dementia and this number is expected to double
by 2040. Despite the number of dementia cases and decades of research, there remains much unknown about
the pathogenesis and progression of the disease, and, at present, no effective treatment exists to arrest or
reverse the cognitive decline associated with the condition. In this context, identification of causal relationships
between modifiable targets and dementia risk is central to the development of evidence-based prevention
strategies and will be critically important in maintaining the long-term health of the ageing public. Blood lipid
levels have been implicated in the aetiology of dementia by genetic linkage and functional cell biology studies,
but current epidemiological evidence has yet to reach a consensus on their role in dementia risk.

>>>>

__Background:__ Identification of causal relationships between modifiable targets and risk of dementia is central to the development of evidence-based prevention strategies. Blood lipid levels have been implicated in the aetiology of dementia by genetic linkage and functional cell biology studies, but current epidemiological evidence has yet to reach a consensus on their role in dementia risk. This thesis sought to triangulate the causal effects of lipid levels (total cholesterol, LDL-c, HDL-c, and triglycerides) on risk of incident dementia (all-cause dementia, Alzheimer's disease, and vascular dementia) using existing and new evidence.

&nbsp;

__Methods:__ Four distinct analyses were conducted. Firstly, a systematic review of 81 published and preprinted studies was used to summarise the existing evidence base. Preprints were searched using a new research tool created as part of this thesis. The existing evidence was then supplemented via two primary studies: i) a cohort study of the association of lipid regulating agents and dementia incidence in the Clinical Practice Research Datalink (CPRD); and ii) an individual patient data meta-analysis of the association of blood lipids with dementia incidence in previously unanalysed cohorts accessed through the Dementia Platform UK. Finally, a novel quantitative triangulation framework was proposed, building on recent developments in risk-of-bias assessment and bias-/indirectness-adjusted meta-analyses. This framework was then used to quantitatively triangulate the evidence identified and produced by this thesis.

&nbsp;

__Results:__ The systematic review did not identify a consistent effect of blood lipids on any dementia outcome across study designs, through there was some suggestion of a protective effect of LDL-c lowering on all-cause dementia and Alzheimer’s disease in observational studies of statin use. The analysis of lipid regulating agents in the CPRD provided weak evidence for a protective effect of statins on all-cause dementia and Alzheimer's disease but suggested a harmful association with vascular dementia. However, the use of control outcomes illustrated this finding was likely due to confounding by indication related to vascular factors. The individual patient data meta-analysis suffered from a low response rate to data access requests, but in the three cohorts analysed, there was some evidence only for the association of triglycerides and vascular dementia. Finally, the triangulation analysis integrating the results of the previous three studies did not provide strong evidence for the causal effect of blood lipids on dementia outcomes.

&nbsp;

__Conclusions:__
This thesis provides new evidence concerning the role of blood lipids as a modifiable risk factor for dementia and highlights the uncertainty that still remains in relation to this causal question. In addition, it has developed new evidence synthesis methods and tools, specifically around the inclusion of preprints in systematic reviews and the quantitative triangulation of evidence sources.

### 2021-12-02 index.Rmd
End of day:

---
#####################
## thesis metadata ##
#####################
title: |
   Lipids and dementia:\
   An investigation of their relationship

>>>>

---
#####################
## thesis metadata ##
#####################
title: |
   Determining the causal effects of lipid levels on risk of dementia: a triangulation of new and existing evidence

### 2021-12-02 manuscripts/cprd/manuscript.Rmd
End of day:

A further limitation stems from uncontrolled confounding due to genetic factors. Number of ApoE $\epsilon4$ alleles represents the strongest genetic risk factor for Alzheimer's disease, but also substantially increases LDL-c levels,[@bennet2007] potentially prompting treatment with a statin or other lipid regulating agent. We were unable to control for ApoE genotype in this analysis as we did not have access to genetic data on participants. As a result, any protective association between LRA use and the Alzheimer's disease outcomes may be masked by residual negative confounding by ApoE.

>>>>

A further limitation stems from uncontrolled confounding due to genetic factors. Number of _ApoE_ $\epsilon4$ alleles represents the strongest genetic risk factor for Alzheimer's disease, but also substantially increases LDL-c levels,[@bennet2007] potentially prompting treatment with a statin or other lipid regulating agent. We were unable to control for _ApoE_ genotype in this analysis as we did not have access to genetic data on participants. As a result, any protective association between LRA use and the Alzheimer's disease outcomes may be masked by residual negative confounding by ApoE.
